---
output:
  html_document:
    keep_md: yes
---
```{r setup, echo = F, message=F}
library(knitr)
opts_chunk$set(fig.path = "images/teamreview2104-")
```
#Team Review 2014 - McLaren 

```{r}
source('ergastR-core.R')
```

```{r}
button=driverResults.df('2014','button')
#button
```

```{r}
magnussen=driverResults.df('2014','kevin_magnussen')
#magnussen
```


```{r}
hamilton=driverResults.df('2014','hamilton')
#hamilton
```

```{r}
rosberg=driverResults.df('2014','rosberg')
#rosberg
```

```{r message=FALSE,warning=FALSE}
library(ggplot2)
newpoints=data.frame(x=c(25,18,15,12,10,8,6,4,2,1,0),y=c(18,15,12,10,8,6,4,2,1,0,0))
raceoff=merge(button[,c('points','position','positionText','round')],magnussen[,c('points','position','positionText','round')],by='round')
raceoff$topd2=(raceoff$position.x>raceoff$position.y)

#Hack final round points
raceoff[raceoff['round']==19,]$points.x=raceoff[raceoff['round']==19,]$points.x/2
raceoff[raceoff['round']==19,]$points.y=raceoff[raceoff['round']==19,]$points.y/2
raceoff=ddply(raceoff, .(round),transform,
              d1pos=max(position.x,position.y),
              d2pos=min(position.x,position.y),
              d1points=max(points.x,points.y),
              d2points=min(points.x,points.y))
  
library(plyr)
raceoff.summary=ddply(raceoff,.(points.x,points.y),summarise,cnt=length(round))
raceoff.team.summary=ddply(raceoff,.(d1points,d2points),summarise,cnt=length(round))
```

```{r}
pointsPerformanceChart=function(raceoff,raceoff.team.summary,constructorRef){
  ggplot(raceoff.team.summary,aes(x=d1points,y=d2points))+stat_smooth(data=raceoff,aes(x=d1points,y=d2points),se=F,col='magenta')+geom_line(data=newpoints,aes(x=x,y=y),col='red',linetype='dotted')+geom_point(aes(size=cnt),col='pink')+geom_text(aes(label=cnt),size=2)+xlab('Highest placed driver points')+ylab('Lower placed driver points') +scale_size_continuous(range=c(3,8))+ggtitle(constructorRef)+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL) +scale_x_continuous(breaks = newpoints$x,minor_breaks=NULL)+coord_fixed()+guides(size=FALSE)+theme_bw()
}
```

```{r mclaren_pointsPerf}
pointsPerformanceChart(raceoff,raceoff.team.summary,'McLaren')
```

```{r}
pointsDualPerformanceChart=function(raceoff,raceoff.summary,title,d1,d2){
  ggplot(raceoff.summary,aes(x=points.x,y=points.y))+stat_smooth(data=raceoff,aes(x=d1points,y=d2points),se=F,col='magenta')+geom_line(data=newpoints,aes(x=x,y=y),col='red',linetype='dotted')+geom_line(data=newpoints,aes(x=rev(y),y=rev(x)),col='blue',linetype='dotted')+stat_abline(col='grey')+geom_point(aes(size=cnt),col='pink')+geom_text(aes(label=cnt),size=2)+scale_size(range=c(2,6),name="Count")+xlab(d1)+ylab(d2)+ggtitle(title)+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL) +scale_x_continuous(breaks = newpoints$x,minor_breaks=NULL)+coord_fixed()+theme_bw()+guides(size=FALSE)
  }
```


"If we are going to make a mark, it may as well be a meaningful one. The simplest - and most useful - meaningful mark is a digit." John W Tukey, "Some Graphic and Semigraphic Displays", commenting on stem-and-leaf plots, in T A Bancroft,ed., Statistical Papers in Honot of George W Snedecor,1972, p296, quoted in Edward R. Tufte, Envisioning Information, 1990, p46.

```{r mclaren_pointsPerfDual}
pointsDualPerformanceChart(raceoff,raceoff.summary,"Points Awarded by Race, McLaren, 2014",'BUT','MAG')
```

```{r}
driverPos=function(raceoff,d1,d2){
  ggplot(raceoff,aes(x=round))+geom_hline(yintercept=3.5,colour='grey',linetype='dotted')+geom_hline(yintercept=10.5,colour='grey',linetype='dashed')+geom_hline(yintercept=0.5,colour='grey',linetype='dotted')+geom_hline(yintercept=22.5,colour='grey',linetype='dotted')+geom_text(aes(y=position.x,fontface=ifelse((positionText.x=='R'), 'italic' , 'plain')),label=d1,size=3)+geom_text(aes(y=position.y,fontface=ifelse((positionText.y=='R'), 'italic' , 'plain')),label=d2,size=3)+xlab('Round')+ylab('Position')+geom_segment(aes(x=round,xend=round,y=d1pos-0.25,yend=d2pos+0.25,col=topd2)) +guides(colour=FALSE)+scale_x_continuous(breaks = 1:22,minor_breaks=NULL)+scale_y_reverse(breaks = c(1,5,10,15,20),minor_breaks=1:22)+theme_bw()
  }

driverPoints=function(raceoff,d1,d2){
  ggplot(raceoff,aes(x=round))+geom_hline(yintercept=0.5,colour='grey',linetype='dotted')+geom_hline(yintercept=24.5,colour='grey',linetype='dotted')+geom_text(aes(y=points.x),label=d1,size=3)+geom_text(aes(y=points.y),label=d2,size=3)+xlab('Round')+ylab('Points')+geom_segment(aes(x=round,xend=round,y=d1points-0.25,yend=d2points+0.25,col=topd2)) +guides(colour=FALSE)+scale_x_continuous(breaks = 1:22,minor_breaks=NULL)+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL,limits=c(0.2,25))+theme_bw()
  }
```

```{r driverPoints_mclaren}
driverPoints(raceoff,'BUT','MAG')
```


If we further annotate these charts with a text label that identifies the qualifying and grid position of each driver, we can provide an even more complete summary of how the drivers compared over the qualifying and race sessions. To maximise the use of space, we can drop the name labels and instead just rely on the colour field to identify which driver was higher placed at the end of the race.

...

Need a 'last classified position' line...
...

```{r driverPos_mclaren}
driverPos(raceoff,'BUT','MAG')
```

                      
```{r}
ggplot(raceoff,aes(x=round))+geom_text(aes(y=points.x),label='BUT',size=3)+geom_text(aes(y=points.y),label='MAG',size=3)+xlab('Round')+ylab('Points')+geom_segment(aes(x=round,xend=round,y=d1points-0.25,yend=d2points+0.25,col=topd2))+ylim(0.2,25)+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL)+guides(colour=FALSE)
```

```{r}
ggplot(raceoff.team.summary,aes(x=d1points,y=d2points))+stat_smooth(se=F,col='magenta')+geom_line(data=newpoints,aes(x=x,y=y),col='red',linetype='dotted')+geom_point(aes(size=cnt))+xlab('Highest placed driver points')+ylab('Lower placed driver points') +scale_size_continuous(range=c(2,6),limits = c(1, 3),breaks = c(1,2,3))+ggtitle('McLaren')+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL) +scale_x_continuous(breaks = newpoints$x,minor_breaks=NULL)+coord_fixed()

```

```{r}
ggplot(raceoff.summary,aes(x=points.x,y=points.y))+stat_smooth(se=F,col='magenta')+geom_line(data=newpoints,aes(x=x,y=y),col='red')+geom_line(data=newpoints,aes(x=rev(y),y=rev(x)),col='blue')+stat_abline(col='grey')+geom_point(aes(size=cnt))+xlab('BUT')+ylab('MAG') +scale_size_continuous(range=c(2,6),limits = c(1, 3), breaks = c(1,2,3),name="Count")+ggtitle("Points awarded, McLaren, 2014")+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL) +scale_x_continuous(breaks = newpoints$x,minor_breaks=NULL)+coord_fixed()

```

```{r warning=FALSE,message=FALSE}
poslist=data.frame(x=1:23,y=2:24)
raceoff.summary=ddply(raceoff,.(position.x,position.y),summarise,cnt=length(round))
raceoff.team.summary=ddply(raceoff,.(d1points,d2points),summarise,cnt=length(round))



ggplot(raceoff.summary,aes(x=position.x,y=position.y))+stat_smooth(se=F,col='magenta')+geom_line(data=poslist,aes(x=x,y=y),col='red')+geom_line(data=poslist,aes(x=y,y=x),col='blue')+stat_abline(col='grey')+geom_point(aes(size=cnt))+scale_size(range=c(2,6))+scale_x_reverse()+scale_y_reverse()+xlab('BUT')+ylab('MAG')+scale_size_continuous(range=c(2,6),limits = c(1, 25), breaks = c(1,2,3))
```

```{r warning=FALSE,message=FALSE}
raceoff=merge(hamilton[,c('points','position','positionText','round')],rosberg[,c('points','position','positionText','round')],by='round')
raceoff$topd2=(raceoff$position.x>raceoff$position.y)
#Hack final round points
raceoff[raceoff['round']==19,]$points.x=raceoff[raceoff['round']==19,]$points.x/2
raceoff[raceoff['round']==19,]$points.y=raceoff[raceoff['round']==19,]$points.y/2
raceoff=ddply(raceoff, .(round),transform,
              d1pos=max(position.x,position.y),
              d2pos=min(position.x,position.y),
              d1points=max(points.x,points.y),
              d2points=min(points.x,points.y))

raceoff.summary=ddply(raceoff,.(points.x,points.y),summarise,cnt=length(round))
raceoff.team.summary=ddply(raceoff,.(d1points,d2points),summarise,cnt=length(round))
```
```{r mercedes_pointsPerf}
pointsPerformanceChart(raceoff,raceoff.team.summary,'Mercedes')
```


```{r}
ggplot(raceoff,aes(x=round))+geom_text(aes(y=points.x),label='HAM',size=3)+geom_text(aes(y=points.y),label='ROS',size=3)+xlab('Round')+ylab('Points')+geom_segment(aes(x=round,xend=round,y=d1points-0.25,yend=d2points+0.25,col=topd2))+scale_y_continuous(breaks = newpoints$x,minor_breaks=NULL) +guides(colour=FALSE)+scale_x_continuous(breaks = 1:22,minor_breaks=NULL)
```


```{r driverPoints_mercedes}
driverPoints(raceoff,'HAM','ROS')
```

```{r driverPos_mercedes}
driverPos(raceoff,'HAM','ROS')
```




```{r mercedes_pointsDualPerf}
pointsDualPerformanceChart(raceoff,raceoff.summary,"Points Awarded by Race, Mercedes, 2014",'HAM','ROS')
```


```{r warning=FALSE,message=FALSE}
raceoff.summary=ddply(raceoff,.(position.x,position.y),summarise,cnt=length(round))

ggplot(raceoff.summary,aes(x=position.x,y=position.y))+stat_smooth(se=F,col='magenta')+geom_line(data=poslist,aes(x=x,y=y),col='red')+geom_line(data=poslist,aes(x=y,y=x),col='blue')+stat_abline(col='grey')+geom_point(aes(size=cnt))+scale_size(range=c(2,6))+scale_x_reverse()+scale_y_reverse()+xlab('HAM')+ylab('ROS')+guides(size=FALSE)
```
