---
output:
  html_document:
    keep_md: yes
---
```{r setup, echo = F, message=F}
require(knitr)
opts_chunk$set(fig.path = "images/movingaverage-")
```
# Consistent Performers? 

moving average - measure of consistency
 


curveball p136


```{r message=FALSE}
require(DBI)
ergastdb =dbConnect(RSQLite::SQLite(), './ergastdb13.sqlite')
#Helper function to display database query result as a formatted table
kdb=function(q){ kable(dbGetQuery(ergastdb,q)) }
```


```{r}
onetwoFlag=dbGetQuery(ergastdb,
           'SELECT r.name,r.year,round,constructorRef,
           CASE WHEN (MAX(res.position)<=2 AND COUNT(res.position)=2) THEN 1 ELSE 0 END AS onetwo
           FROM results res JOIN races r JOIN constructors c
           WHERE r.raceId=res.raceId
           AND c.constructorId=res.constructorId 
           AND constructorRef="red_bull" 
           GROUP BY res.raceId,res.constructorid 
           ORDER BY r.year, round')

kable(head(onetwoFlag))
```

```{r}
doublepodiumFlag = dbGetQuery(ergastdb,
           'SELECT DISTINCT r.name, year, round, constructorRef, 
              CASE WHEN (COUNT(res.position)=2 and MAX(res.position)<=3) 
                THEN 1 ELSE 0 END AS doublePodium 
           FROM results res JOIN races r JOIN constructors c 
           WHERE r.raceId=res.raceId 
           AND c.constructorId=res.constructorId 
           AND constructorRef="red_bull" 
           GROUP BY res.raceId,res.constructorid 
           ORDER BY year, round')
```           
```{r}
#y contains a list of result flags
#val defines polarity
streaks=function(y,val=0){
  #Start by initialising run length to 0
  run=0
  #Define a list to capture all the runs, in sequence
  runs=c()
  #Initialise a variable that contains the previous result
  prev=y[1]
  #The last flag identifies the last result as part of a run
  last=TRUE
  #Search through each result flag
  for (i in y) {
    #Is the current result is the same as the previous one?
    if (i!=prev) {
      #If not, record the length of the previous run, and its polarity
      runs=c(runs,run*(if (prev==val) -1 else 1))
      #Initialise the next run length
      run=0
      #This result is the first in a new run
      last=FALSE
    } else {
      #We are still in a run
      last=TRUE
    }
    #Keep track of what the previous result flag was
    prev=i
    #Increase the length of the run counter
    run=run+1
  }
  #If the last result was part of a run, record that run
  if (last | (run==1)) runs=c(runs,run*(if (prev==val) -1 else 1))
  #Create a dataframe from run list
  ss=data.frame(l=runs)
  #Tally how many results in total have been counted after each run
  #That is, record the result number for the last result in each run
  ss$end=cumsum(abs(ss$l))
  #Identify the result number for the start of each run
  ss$start=ss$end-abs(ss$l)+1
  #Reorder the columns
  ss[,c("start","end","l")]
}
```


```{r}
streakReview=function(driverRef,years='',length=5,typ=2,mode='podium',topN=''){
  #The definitions are a bit messy...
  #If we set a topN parameter, force the topN mode
  if (topN!='') mode='topN'
  
  #Get the driver details
  driver=getDriver(driverRef,years)
  
  #Calculate the desired 0/1 stream based on the mode property
  if (mode=='podium') driver$mode=mapply(podium,driver$position)
  else if (mode=='frontrow') driver$mode=mapply(frontrow,driver$grid)
  else if (mode=='topN') driver$mode=mapply(topNfinish,driver$position,topN)
  else if (mode=='unclassified') driver$mode=mapply(unclassified,driver$position)
  else if (mode=='inpoints') driver$mode=mapply(inpoints,driver$points)
  else return(data.frame())
  
  #Calculate the streaks in the desired mode property
  streak=streaks(driver$mode)
  
  #Annotate the streak with start (first), end (last) and broken-by information
  streak$startc= mapply(function(x) driver[x,'name'],streak$start) 
  streak$endc= mapply(function(x) driver[x,'name'],streak$end) 
  streak$starty= mapply(function(x) driver[x,'year'],streak$start) 
  streak$endy= mapply(function(x) driver[x,'year'],streak$end)
  nd=nrow(driver)
  streak$brokenbyy=mapply(function(x) if (nd<x+1) NA else driver[x+1,'year'],
                          streak$end)
  streak$brokenbyc= mapply(function(x) if (nd<x+1) NA else driver[x+1,'name'],
                           streak$end)
  
  #The typ argument lets us get all streaks, 1s streaks, or 0s streaks
  #greater than or equal to a specified length 
  if (typ==2) streak[abs(streak["l"])>=length,]
  else if (typ==1) streak[abs(streak["l"])>=length & streak["l"]>0,]
  else streak[abs(streak["l"])>=length & streak["l"]<0,]
}
```

stre

require(zoo)
alonso=getDriver('alonso')
#If a driver isn't classified, set their position at 25 as a penalty
alonso$position[is.na(alonso$position)] = 25
rollmean(alonso$position,3)

#We can also do rolling means for the points
rollmean(vettel$points,3)



#ma=function(x,n=5){filter(x,rep(1/n,n), sides=1)}
##sides=1 past only, sides=2 past and future
# ma(alonso$position)
 ```
 