---
output:
  html_document:
    keep_md: yes
---
```{r setup, echo = F, message=F}
library(knitr)
opts_chunk$set(fig.path = "images/weekend-")
```

```{r}
source("practiceQualiLaps.R")
```

```{r}
qlaps= read.csv("~/Dropbox/various/aus17_qualilaptimes.csv")

```

quali_utilisation
```{r quali_utilisation,fig.width=10}
qlapsa=rawLap_augment_laptimes(qlaps)
augmented_session_utilisation_chart(qlapsa,3)
```

Re: the below - need to go from official timing sheets - website is random junk...
```{r}
source('fia_classification_scraper.R')
source('qualicharts.R')
##Session classifications page
xx=pageGrabber('http://www.fia.com/events/fia-formula-one-world-championship/season-2017/session-classifications-19')
```

quali_slopegraph
```{r quali_slopegraph}
qr=quali_progression_ergast_tx(ergastifyQualiClass(fiaSessionClassQualifyingTidy(fiaTableGrabber(xx,4))))
qrm=quali_progression_ergast_melt(qr)
core_qualifying_rank_slopegraph(qr,qrm,spacer=0.21)
```

#?following broken becuase one of runners is missing times in final session?
quali_slopegraph_time
```{r quali_slopegraph_time}
core_qualifying_time_slopegraph(qr,qrm,spacer=0.21)+ylim(92,96)
```

quali_utilisationb
```{r quali_utilisationb,fig.width=10}
source('qualicharts.R')
qlapsb=rawLap_augment_quali(qlaps)
#qlapsb=qsessionOverride(qlapsb,0,2000,3100)
qlapsb=quali_purplePatch(qlapsb)
g=plot_session_utilisation_chart(qlapsb,3)

qlapsc= read.csv("~/Dropbox/various/mon16_racePrelimClass.csv")
qlapsc$pos=as.integer(qlapsc$pos)
levels(qlapsc$name) =qlapsc$name[order(qlapsc$pos)]
g=g+scale_y_discrete(limits=rev(levels(qlapsc$name)))
g
```

TEST
```{r}
library(grid)
library(gridExtra)
grid.arrange(g, nrow=1,
             #top="My title", 
             bottom = textGrob("To learn how to create this chart: f1datajunkie.com",
                               gp = gpar(fontface=3, fontsize=9),
                               hjust=1, x=1))
```

```{r}
qlapsb
```

qualisessionplot
```{r}
#qlapsb=quali_purplePatch(qlapsb)
```
Not sure if the above works - eg where one of druvers does not run in final session

The 13 was red flagged and auto session numbering mistakenly counted four sessions - need to remap the sessions

```{r}
print(qlapsb)
```

Following may be broken in situation where one of drivers does not run in final session - or another problem?

```{r qualisessionplot}
#TO DO
#The rank ordering and gap calculations need changing for qualifying
quali_session_times_plots(qlapsb)
```

quali1_cutoff
```{r quali1_cutoff}
quali_session_times_plots(qlapsb[qlapsb['qsession']==1,],1,FALSE,FALSE,TRUE,c(15,10,2))+theme_bw()
```

quali2_cutoff
```{r quali2_cutoff}
quali_session_times_plots(qlapsb[qlapsb['qsession']==2,],2,FALSE,FALSE,TRUE,c(15,10,2))+theme_bw()
```

quali3_cutoff
```{r quali3_cutoff}
quali_session_times_plots(qlapsb[qlapsb['qsession']>2,],3,FALSE,FALSE,TRUE,c(15,10,2))+theme_bw()
```



```{r}
#qr=qualiResults.df(2015,2)
#qr=quali_progression_ergast_tx(qr)
#qrm=quali_progression_ergast_melt(qr)
#core_qualifying_rank_slopegraph(qr,qrm,spacer=0.21)
```

```{r}
source("code_battlemap.R")
lapTimes=lapsData.df(2016,6)
lapTimes=battlemap_encoder(lapTimes)
```

```{r aut_racechart}
#Grab some data
#lapTimes =lapsData.df(2015,7)
 
#Process the laptimes
#lapTimes=battlemap_encoder(lapTimes)
 
#Find the accumulated race time at the start of each leader's lap
lapTimes=ddply(lapTimes,.(leadlap),transform,lstart=min(acctime))
 
#Find the on-track gap to leader
lapTimes['trackdiff']=lapTimes['acctime']-lapTimes['lstart']
 
#Construct a dataframe that contains the difference between the 
#leader accumulated laptime on current lap and next lap
#i.e. how far behind current lap leader is next-lap leader?
ll=data.frame(t=diff(lapTimes[lapTimes['position']==1,'acctime']))
#Generate a de facto lap count
ll['n']=1:nrow(ll)
#Grab the code of the lap leader on the next lap
ll['c']=lapTimes[lapTimes['position']==1 & lapTimes['lap']>1,'code']
 
#Plot the on-track gap to leader versus leader lap
g = ggplot(lapTimes) 
g=g+geom_text(data=lapTimes[lapTimes['position']==1,],aes(x=-3,y=leadlap,label=code),size=2)
g = g + geom_point(aes(x=trackdiff,y=leadlap,col=(lap==leadlap)), pch=1)
g = g + geom_point(data=lapTimes[lapTimes['driverId']=='bottas',],
                  aes(x=trackdiff,y=leadlap), pch='+')
g = g + geom_text(data=lapTimes[lapTimes['lapsbehind']>0,],
                  aes(x=trackdiff,y=leadlap, label=lapsbehind),size=3)
g = g + geom_point(data=ll,aes(x=t, y=n), pch='x')
g = g + geom_text(data=ll,aes(x=t+3, y=n,label=c), size=2)
g = g + geom_vline(aes(xintercept=19), linetype=3)
g+xlim(-4,125)+ guides(colour=FALSE)
```

Find purple times:


lapTimes=arrange(lapTimes,acctime)
#Purple time can't be on an inlap or an outlap
#?? and outlap is the lap after  pitstop:
p=pitsData.df(2015,9)
#merge pitdata with lapsdata
lapTimesp=merge(lapTimes,p,by = c('lap','driverId'),all.x=T)
#flag pit laps
lapTimesp$ps = ifelse(is.na(lapTimesp$milliseconds), F, T)
#do an offset on the laps that are pitstops for each driver
lapTimesp=ddply(lapTimesp, .(driverId),transform,outlap=c(FALSE, head(ps,-1)))

#identify lap before pit lap
lapTimesp=arrange(lapTimesp,-acctime)
#lapTimesp['prelap']=c(FALSE, diff(lapTimesp[['ps']])==-1)
lapTimesp=ddply(lapTimesp, .(driverId),transform,prelap=c(FALSE, head(ps,-1)))
lapTimesp=arrange(lapTimesp,acctime)

library(sqldf)
#find undercut

d1 outlap on lap X, d2 outlap on lap X+1
d1 prelap on lap X-2, d1 !=d2
d1 pos on X-2 > d2 pos on X-2
d1 pos on X+1 < d2 pos on X+1

#donlt use this?
ss=sqldf('select d1a.driverId,d2a.driverId,d1a.lap,d1a.position,d2a.position, d1b.lap,d1b.position,d2b.position from lapTimesp d1a, lapTimesp d1b, lapTimesp d2a, lapTimesp d2b where d1a.driverId=d1b.driverId and d2a.driverId=d2b.driverId and d1a.driverId!=d2a.driverId and d1a.prelap and d1a.lap=d2a.lap and d2b.outlap and d2b.lap=d1b.lap and d1a.lap=d1b.lap-3 and d1a.position>d2a.position and d1b.position < d2b.position')


#use this?
#This one gives us a window of laps within which the undercut must take place
#the 2 outlap lap is at least 3 more than the 1 prelap (0:prelap/., 1:ps/.,2:outlap/ps 3:./outlap)
ss=sqldf('select d1a.driverId,d2a.driverId,d1a.lap,d1a.position,d2a.position, d1b.lap,d1b.position,d2b.position from lapTimesp d1a, lapTimesp d1b, lapTimesp d2a, lapTimesp d2b where d1a.driverId=d1b.driverId and d2a.driverId=d2b.driverId and d1a.driverId!=d2a.driverId and d1a.prelap and d1a.lap=d2a.lap and d2b.outlap and d2b.lap=d1b.lap and (d1a.lap+3<=d1b.lap and d1b.lap<=d1a.lap+2+5) and d1a.position>d2a.position and d1b.position < d2b.position')

#also other strategy of staying out longer time from behind and then coming out in front?

lapTimes['ptime']=cummin(lapTimes['rawtime']+900*lapTimes['pitflag']++900*lapTimes['outlapflag'])
lapTimes['purple']=(lapTimes['ptime']==lapTimes['rawtime'])
g+ geom_point(data=lapTimes[lapTimes['ptime']==TRUE,],aes(x=trackdiff,y=leadlap),colour='purple', pch=1)

```{r}
g = ggplot(lapTimes)
g=g+geom_text(data=lapTimes[lapTimes['position']==1,],aes(x=-1,y=leadlap,label=code),size=2)
g = g + geom_point(aes(x=trackdiff,y=leadlap,col=(lap==leadlap)), pch=1)
g = g + geom_point(data=lapTimes[lapTimes['driverId']=='vettel',],
                  aes(x=trackdiff,y=leadlap), pch='+')
g = g + geom_text(data=lapTimes[lapTimes['lapsbehind']>0,],
                  aes(x=trackdiff,y=leadlap, label=lapsbehind),size=3)
g = g + geom_vline(aes(xintercept=19), linetype=3)
g+xlim(-3,25)+ guides(colour=FALSE)
```
BOT
```{r battle_aut_BOT}
battlemapFull_byDriver(lapTimes,'BOT',"2016 Monaco - BOT's Race",c(-20,20))
```

VET
```{r battle_aut_VET}
battlemapFull_byDriver(lapTimes,'VET',"2016 Monaco - VET's Race",c(-25,10))
```

ROS
```{r battle_aut_ROS}
battlemapFull_byDriver(lapTimes,'ROS',"2016 Monaco - ROS's Race",c(-5,15))
```


VES
```{r battle_aut_VES}
battlemapFull_byDriver(lapTimes,'VES',"2016 Monaco - VES's Race",c(-10,10))
```


PER
```{r battle_aut_PER}
battlemapFull_byDriver(lapTimes,'PER',"2016 Monaco - PER's Race",c(-10,20))
```

SAI
```{r battle_aut_SAI}
battlemapFull_byDriver(lapTimes,'SAI',"2016 Monaco - SAI's Race",c(-15,5))
```

HAM
```{r battle_aut_HAM}
battlemapFull_byDriver(lapTimes,'HAM',"2016 Monaco - HAM's Race",c(-20,5))
```

MAS
```{r battle_aut_MAS}
battlemapFull_byDriver(lapTimes,'MAS',"2016 Monaco - MAS's Race",c(-20,20))
```

```{r battle_aut_RAI}
battlemapFull_byDriver(lapTimes,'RAI',"2016 Monaco - RAI's Race",c(-20,20))
```

```{r battle_aut_MAG}
battlemapFull_byDriver(lapTimes,'MAG',"2016 Monaco - MAG's Race",c(-20,20))
```

```{r battle_aut_BUT}
battlemapFull_byDriver(lapTimes,'BUT',"2016 Monaco - BUT's Race",c(-15,10))
```


```{r battle_aut_WEH}
battlemapFull_byDriver(lapTimes,'WEH',"2016 Monaco -WEH's Race",c(-10,20))
```

```{r battle_aut_GUT}
battlemapFull_byDriver(lapTimes,'GUT',"2016 Monaco -GUT's Race",c(-15,20))
```

```{r battle_aut_GRO}
battlemapFull_byDriver(lapTimes,'GRO',"2016 Monaco - GRO's Race",c(-20,10))
```

```{r battle_aut_ALO}
battlemapFull_byDriver(lapTimes,'ALO',"2016 Monaco - ALO's Race",c(-10,30))
```



```{r battle_aut_RIC}
battlemapFull_byDriver(lapTimes,'RIC',"2016 Monaco - RIC's Race",c(-20,10))
```

```{r battle_aut_PAL}
battlemapFull_byDriver(lapTimes,'PAL',"2016 Monaco - PAL's Race",c(-26,20))
```

```{r battle_aut_HUL}
battlemapFull_byDriver(lapTimes,'HUL',"2016 Monaco - HUL's Race",c(-5,10))
```

```{r battle_aut_ERI}
battlemapFull_byDriver(lapTimes,'ERI',"2016 Monaco - ERI's Race",c(-20,20))
```

```{r battle_aut_VAN}
battlemapFull_byDriver(lapTimes,'KVY',"2016 Monaco - VAN's Race",c(-15,20))
```

